A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-20 docs/paper.tex
Typos and language fixes

In order to attempt to improve the performance of the python language as a whole we plan to attempt to improve the performance of python dictionaries.

@@

One of the features that we added to the python source was a conditional flag
that allowed us to view dictionary statistics (collisions, probes, accesses,
etc.) for a given invocation of the Python binary.  With that modification we
were able to see that this code used 92 lookups in string dictionaries and 24
lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
hits in a 3 line piece of code which never uses dictionaries explicitly is quite
significant.  We explain this extensive dictionary usage below.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we mostly limit ourselves to the study of new implementations of
hashing the string type for the rest of the paper.

@@

Linear probing is appealing because of its excellent cache behavior in comparison to the probing scheme currently utilized in Python.  Consider the Intel Core 2 Duo processor with a 64 byte cache line.  Each Python dictionary entry is 12 bytes so when Python’s current probing scheme probes another slot, it is not in the current cache line so it has to pull another cache line in from memory, but with linear probing, that next slot is most likely in the current cache line.

However, Patrascu et al. [2] describe simple tabulation hashing for fixed key widths.  For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).  Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how to implement this. In the following section we explain the implementation details of our approach to these problems.

\subsection{Implementation Details}
We prepopulated random tables in memory with random bits from random.org, which
sources randomness form atmospheric noise in order to remove the reliance on
pseudo-random number generators.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing it in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to do this.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy for fewer tables and too much memory bandwidth for a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the
valuable benchmarks is to see how the new dictionary performs on them. We used
PyBench benchmark suit for this task. We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measures for running the whole benchmark suite.

\begin{table}
  \caption{Performance results on selected components of the PyBench suite.  We
    have omitted the least significant performance results in terms of percent
    change.
  }
\begin{tabular}{| l | c |}
  \hline
  \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
  NormalClassAttribute &  19.20\% \\ \hline
  SpecialClassAttribute &  18.88\% \\ \hline
  NormalInstanceAttribute &  17.77\% \\ \hline
  StringPredicates &  15.25\% \\ \hline
  UnicodePredicates &  13.82\% \\ \hline
  SimpleListManipulation &  13.03\% \\ \hline
  CompareUnicode &  11.27\% \\ \hline
  SimpleComplexArithmetic &  10.83\% \\ \hline
  SpecialInstanceAttribute &  10.52\% \\ \hline
  TryFinally &  10.35\% \\ \hline
  CreateUnicodeWithConcat &  8.52\% \\ \hline
  StringSlicing &  6.52\% \\ \hline
  Small Performance Changes Omitted & ... \\ \hline
  UnicodeMappings &  -1.27\% \\ \hline
  ListSlicing &  -2.40\% \\ \hline
  ComplexPythonFunctionCalls &  -3.55\% \\ \hline
  TupleSlicing &  -3.72\% \\ \hline
  WithRaiseExcept &  -5.89\% \\ \hline
  TryRaiseExcept &  -11.99\% \\ \hline
  Totals & 4.54\% \\ \hline
\end{tabular}
\end{table}

PyBench showed the performance difference in each of the individual CPython internals however for a more realistic picture of the performance in real world programs more extensive benchmarking of the new implementation was needed. Especially since we wanted to know the frequency of the different CPython internals in real-world applications and therefore the corresponding runtime changes. The new implementation was tested on several standard real world programs such as Tornado web server, Django web framework, html5lib and nearly 10 other real-world programs.

@@

Different request rates (up to 2000 requests per second) were tested on this
sample server and their corresponding response rates were calculated using
httperf, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates. This finding followed by other benchmarks showed that the new implementation gives more consistent runtime on different iterations of a same code compared to the current CPython implementation. This consistency is a desired behavior. We suspect that this behavior is a result of the randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

As mentioned before, another set of the benchmarks that we utilized were based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Percentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
\end{figure}

@@

\section{Conclusion and future work}
Throughout the course of our work we came to the following conclusions:
\begin{enumerate}
\item Hardware specific performance issues dominate at small scales: When you are attempting to work on a process that runs as fast as a hash table lookup or insertion already does, there is little room for error.  Improving hash performance with simple tabulation is not an easy task simply because any change you make will probably end up making the hash table run slower.  That said:

@@

In terms of future work, we have 2 avenues of attack. First, we could use different dictionaries for different purposes. Many of the python dictionaries are small, and their keysets never change. No one is adding a new built-in function in the near future. 
Improve performance bottlenecks illustrated by benchmarks

>>>>

In order to attempt to improve the performance of the Python language as a whole we plan to attempt to improve the performance of Python dictionaries.

@@

One of the features that we added to the Python source was a conditional flag
that allowed us to view dictionary statistics (collisions, probes, accesses,
etc.) for a given invocation of the Python binary.  With that modification we
were able to see that this code used 92 lookups in string dictionaries and 24
lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
hits in a 3 line piece of code which never uses dictionaries explicitly is quite
significant.  We explain this extensive dictionary usage below.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we predominantly focus on the study of new implementations of
hashes for the string type for the rest of the paper.

@@

Linear probing is appealing because of its excellent cache behavior in
comparison to the probing scheme currently utilized in Python.  Consider the
Intel Core 2 Duo processor with a 64 byte cache line.  Each Python dictionary
entry is 12 bytes so when Python probes a dictionary slot, it pulls in at least
4 more dictionary entries with it.  However, when the current probing scheme
probes another slot, it is not in the current cache line so it has to pull
another cache line in from memory.  With linear probing, that next slot is most likely in the current cache line.

Patrascu et al. [2] describe simple tabulation hashing for fixed key widths.  For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).  Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how to implement this. In the following section we explain the implementation details of our approach to these problems.

\subsection{Implementation Details}
We modified the source code of CPython 2.7.3 to use simple tabulation hashing
with linear probing.  We use a fixed number of tabulation tables and cycle between them
for variable length strings.

We prepopulated random tables in memory with random bits from random.org, which
sources randomness from atmospheric noise in order to remove the reliance on
pseudo-random number generators.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing the tabulation tables in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to generate these instructions.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted using 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the
valuable benchmarks is to see how the new dictionary performs on them. We used
PyBench benchmark suit for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

\begin{table}
  \caption{Performance results on selected components of the PyBench suite.  We
    have omitted the least significant performance results in terms of percent
    change.
  }
\begin{tabular}{| l | c |}
  \hline
  \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
  NormalClassAttribute &  19.20\% \\ \hline
  SpecialClassAttribute &  18.88\% \\ \hline
  NormalInstanceAttribute &  17.77\% \\ \hline
  StringPredicates &  15.25\% \\ \hline
  UnicodePredicates &  13.82\% \\ \hline
  SimpleListManipulation &  13.03\% \\ \hline
  CompareUnicode &  11.27\% \\ \hline
  SimpleComplexArithmetic &  10.83\% \\ \hline
  SpecialInstanceAttribute &  10.52\% \\ \hline
  TryFinally &  10.35\% \\ \hline
  CreateUnicodeWithConcat &  8.52\% \\ \hline
  StringSlicing &  6.52\% \\ \hline
  Small Performance Changes Omitted & ... \\ \hline
  UnicodeMappings &  -1.27\% \\ \hline
  ListSlicing &  -2.40\% \\ \hline
  ComplexPythonFunctionCalls &  -3.55\% \\ \hline
  TupleSlicing &  -3.72\% \\ \hline
  WithRaiseExcept &  -5.89\% \\ \hline
  TryRaiseExcept &  -11.99\% \\ \hline
  Totals & 4.54\% \\ \hline
\end{tabular}
\label{pybenchtable}
\end{table}

PyBench showed the performance difference in each of the individual CPython
internals however for a more realistic picture of the performance in real world
programs more extensive benchmarking of the new implementation was needed.
Especially since we wanted to know the frequency of the different CPython
internals in real-world applications and therefore the corresponding runtime
changes. The new implementation was tested on several standard real world
programs such as Tornado web server, Django web framework, html5lib and ten other real-world programs.

@@

Different request rates (up to 2000 requests per second) were tested on this
sample server and their corresponding response rates were calculated using
httperf, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates.

This finding as well as those by other benchmarks showed that the new
implementation gives more consistent runtime on different iterations of a same
code compared to the current CPython implementation. This consistency is a
desired behavior. We suspect that this behavior is a result of the "better" randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

As mentioned before, another set of the benchmarks that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Percentage of overall improvements in some real-world Python
  applications. Nearly none of the applications performed worse. Python Pickle
  and html5lib faster around 3\% to 4\%. }
\end{figure}

@@

\section{Conclusion and future work}
Throughout the course of our work we came to the following conclusions:
\begin{enumerate}
\item Hardware specific performance issues dominate at small scales: When you
are attempting to work on a process that runs as fast as a hash table lookup or
insertion already does, there is little room for error.  Improving hash
performance with simple tabulation is not an easy task simply because any
change you make will probably end up making the hash function run slower.  That said:

@@

In terms of future work, we have 2 avenues of attack. First, we could use different dictionaries for different purposes. Many of the Python dictionaries are small, and their keysets never change. No one is adding a new built-in function in the near future.

Improve performance bottlenecks illustrated by benchmarks
